---
title: "multilevel modeling on pitch data"
author: "Leigha DeRango"
date: "2023-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# data found on kaggle at: https://www.kaggle.com/datasets/pschale/mlb-pitch-data-20152018?select=pitches.csv
```

#libraries
```{r}
library(dplyr)
library(ggplot2)
library(ggvis)
library(gridExtra)
library(leaps)
library(lme4)
library(purrr)
```


# join on ab_id
```{r}

joined = full_join(pitches, atbats, by = 'ab_id') %>%
  select(g_id, ab_id, inning, top, pitcher_id, p_throws, batter_id, stand, outs, b_count, s_count, on_1b, on_2b, on_3b, p_score, b_score)

```


# remove missing
```{r}

# rows with any NAs
missing = joined[!complete.cases(joined),]
# - every row that has any NA is missing every game-state relevant column and there's no other trend in what seems to be causing the information to be gone so they'll all simply be removed

joined = na.omit(joined)

```


# aggregate pitcher's pitch count
```{r}

joined = joined %>% group_by(pitcher_id, g_id) %>%
  mutate(pitch_count = cumsum(outs <= 2)-1) %>%
  relocate(pitch_count, .before = batter_id) 

ab_pitches = joined %>% group_by(ab_id) %>%
  mutate(end_pc = max(pitch_count) + 1, ab_pc = max(pitch_count) - min(pitch_count) + 1) %>%
  relocate(ab_pc, .after = pitch_count) %>%
  relocate(end_pc, .after = ab_pc)

```


# base and game states
```{r}

# return one row to reflect game at beginning of each at bat
atbatstry = ab_pitches %>%
  distinct(pitcher_id, batter_id, inning, outs, .keep_all = TRUE)

# include base_state variable (runners on base)
basestates = atbatstry %>% mutate(base_state = case_when(
                                                       on_1b == 0 & on_2b == 0 & on_3b == 0 ~ '000',
                                                       on_1b == 1 & on_2b == 0 & on_3b == 0 ~ '100', 
                                                       on_1b == 0 & on_2b == 1 & on_3b == 0 ~ '020',
                                                       on_1b == 0 & on_2b == 0 & on_3b == 1 ~ '003',
                                                       on_1b == 1 & on_2b == 1 & on_3b == 0 ~ '120',
                                                       on_1b == 0 & on_2b == 1 & on_3b == 1 ~ '023',
                                                       on_1b == 1 & on_2b == 0 & on_3b == 1 ~ '103',
                                                       on_1b == 1 & on_2b == 1 & on_3b == 1 ~ '123'))

# include game_state variable (outs and runners on base)
basestates$game_state = paste(basestates$outs, basestates$base_state, sep = '')

# add expected runs
expruns = basestates %>% mutate(exp_runs = case_when(
                                                      game_state == '0000' ~ 0.53,
                                                      game_state == '0100' ~ 0.94,
                                                      game_state == '0020' ~ 1.17, 
                                                      game_state == '0003' ~ 1.43,
                                                      game_state == '0120' ~ 1.55,
                                                      game_state == '0103' ~ 1.80,
                                                      game_state == '0023' ~ 2.04,
                                                      game_state == '0123' ~ 2.32,
                                                      game_state == '1000' ~ 0.29,
                                                      game_state == '1100' ~ 0.56,
                                                      game_state == '1020' ~ 0.72, 
                                                      game_state == '1003' ~ 1.00,
                                                      game_state == '1120' ~ 1.00,
                                                      game_state == '1103' ~ 1.23,
                                                      game_state == '1023' ~ 1.42,
                                                      game_state == '1123' ~ 1.63,
                                                      game_state == '2000' ~ 0.11,
                                                      game_state == '2100' ~ 0.24,
                                                      game_state == '2020' ~ 0.33, 
                                                      game_state == '2003' ~ 0.38,
                                                      game_state == '2120' ~ 0.46,
                                                      game_state == '2103' ~ 0.54,
                                                      game_state == '2023' ~ 0.60,
                                                      game_state == '2123' ~ 0.77
                                                      ))

```

# change in expected runs
```{r}

exprunsdiff = expruns %>% group_by(g_id, inning, top) %>%
  mutate(exp_runs_change = lead(exp_runs, default = NA) - exp_runs) %>% 
  mutate(exp_runs_change = ifelse(is.na(exp_runs_change), exp_runs*-1, exp_runs_change))

```

# times faced each batter per game
```{r}

battersfaced = exprunsdiff %>% group_by(g_id, pitcher_id, batter_id) %>%
  mutate(times_faced = cumsum(outs <= 2), .after = pitch_count)

```

# adjust classes
```{r}

battersfaced$g_id = as.factor(battersfaced$g_id)
battersfaced$ab_id = as.factor(battersfaced$ab_id)
battersfaced$top = as.factor(battersfaced$top)
battersfaced$pitcher_id = as.factor(as.character(battersfaced$pitcher_id))
battersfaced$p_throws = as.factor(battersfaced$p_throws)
battersfaced$batter_id = as.factor(battersfaced$batter_id)
battersfaced$base_state = as.factor(battersfaced$base_state)
battersfaced$game_state = as.factor(battersfaced$game_state)
battersfaced$stand = as.factor(battersfaced$stand)
battersfaced$inning = as.integer(battersfaced$inning)

```

# reduce to only starting pitchers
```{r}

battersfaced$pitcher_game = paste(battersfaced$pitcher_id, battersfaced$g_id, sep = '') # add combined variable

firstinnings = battersfaced %>% filter(inning == 1) # select only first innings
firstvec = as.vector(firstinnings$pitcher_game) # make vector

starters = battersfaced %>% filter(pitcher_game %in% firstvec) # filter to only starters

```

# separate relievers frame
```{r}

relievers = battersfaced %>% filter(!pitcher_game %in% firstvec) # filter to only relievers

```

# lag ab_pc
```{r}

lagged_ab = starters %>% group_by(g_id, inning, top) %>% mutate(prev_ab_pc = lag(ab_pc, n = 1, default = 0))

lagged_ab_rel = relievers %>% group_by(g_id, inning, top) %>% mutate(prev_ab_pc = lag(ab_pc, n = 1, default = 0))

```

# base data
```{r}

saber1 = lagged_ab %>% select(g_id, ab_id, inning, top, pitcher_id, p_throws, batter_id, stand,
                             times_faced, pitch_count, prev_ab_pc, p_score, b_score,
                             game_state, exp_runs, exp_runs_change)

saber2 = lagged_ab_rel %>% select(g_id, ab_id, inning, top, pitcher_id, p_throws, batter_id, stand,
                             times_faced, pitch_count, prev_ab_pc, p_score, b_score,
                             game_state, exp_runs, exp_runs_change)

```

# summary visualizations
```{r}

# distribution of game_state
ggplot(saber1, aes(x = game_state)) + geom_histogram(stat = 'count') + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# distribution of change in expected runs
ggplot(battersfaced, aes(x = exp_runs_change)) + 
  geom_histogram() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab('Change in Expected Runs') + 
  ylab('Frequency')


# right- vs. left-handed pitchers
ggplot(battersfaced, aes(x = p_throws)) + geom_histogram(stat = 'count')


# trend of changes in run expectancy as pitch count increases
runchanges = battersfaced %>%
  group_by(pitch_count) %>%
  summarise(exp_runs_change = mean(exp_runs_change))

ggplot(runchanges, aes(x = pitch_count, y = exp_runs_change)) + 
  geom_line() + 
  xlab('Pitch Count') + 
  ylab('Average Change in Run Expectancy')


# trend of changes in run expectancy as batters faced increases
batchanges = battersfaced %>%
  group_by(times_faced) %>%
  summarise(exp_runs_change = mean(exp_runs_change))

ggplot(batchanges, aes(x = times_faced, y = exp_runs_change)) + 
  geom_line() + 
  xlab('Times Faced') + 
  ylab('Average Change in Expected Runs')


# comparison of two drastically different pitchers
subset = saber1 %>% filter(pitcher_id %in% c(519141, 453286))
ggplot(data = subset, aes(x = pitch_count, y=exp_runs_change, group=pitcher_id))+
  geom_smooth(method = "lm", se = TRUE,aes(color=pitcher_id))+
  xlab("Pitch Count")+ylab("Change in Run Expectancy")+
  theme_bw() +
  scale_color_hue(labels=c('Max Scherzer','Drew Pomeranz'))
# 453286 - Max Scherzer
# 519141 - Drew Pomeranz

```

# level 1 exploratory
```{r}

saber1$inning = as.numeric(saber1$inning)
cor(saber1$pitch_count, saber1$inning) # 0.924
  # making sure things make sense

cor(saber1$times_faced, saber1$exp_runs_change) # -0.002

cor(saber1$p_score, saber1$exp_runs_change) # 0.0116

cor(saber1$b_score, saber1$exp_runs_change) # -0.0166

cor(saber1$ab_pc, saber1$exp_runs_change) # 0.0557

cor(saber1$pitch_count, saber1$exp_runs_change) # -0.0266

cor(saber1$ab_pc, saber1$pitch_count) # -0.0375

```


# level 2 exploratory
```{r}

# total number of starters' plate apperances
appearances = table(saber1$pitcher_id) %>% data.frame() %>% arrange(Freq)


# side by side comparisons
subset1 = saber1[saber1$pitcher_id %in% c("458681", "519242", "433587", "450308"),]
ggplot(data= subset1, aes(x=pitch_count,y=exp_runs_change)) +
  geom_point() +
  stat_smooth(method="lm", fullrange=TRUE) +
  xlab("Pitch Count") + ylab("Change in Expected Runs") + 
  facet_wrap( ~ pitcher_id) +
  theme(axis.title=element_text(size=16),
        axis.text=element_text(size=14),
        strip.text=element_text(size=14))

```


# linear models
```{r}

naive.1 = lm(exp_runs_change ~ pitch_count, data = saber1)

summary(naive.1)

ggplot(data=saber1, aes(pitch_count, exp_runs_change)) + 
  geom_line() + 
  geom_smooth(method=  'lm') + 
  xlab('Pitch Count') + 
  ylab('Change in Expected Runs')

plot(naive.1)

```

```{r}

naive.2 = lm(exp_runs_change ~ inning + top + p_throws + stand + times_faced + pitch_count + prev_ab_pc + p_score + b_score + p_score:b_score + game_state, data = saber1)
summary(naive.2)


naive.3 = lm(exp_runs_change ~ inning + top + p_throws + stand + times_faced + pitch_count + game_state, data = saber1)
summary(naive.3)

```

```{r}
# relievers

naive.4 = lm(exp_runs_change ~ pitch_count, data = saber2)

summary(naive.4)

ggplot(data=saber2, aes(pitch_count, exp_runs_change)) + 
  geom_line() + 
  geom_smooth(method=  'lm') + 
  xlab('Pitch Count') + 
  ylab('Change in Expected Runs')

plot(naive.4)

```

```{r}
# relievers w covariates

naive.5 = lm(exp_runs_change ~ inning + top + p_throws + stand + times_faced + pitch_count + prev_ab_pc + p_score + b_score + p_score:b_score + game_state, data = saber2)
summary(naive.5)

naive.6 = lm(exp_runs_change ~ inning + top + p_throws + stand + times_faced + pitch_count + game_state, data = saber2)
summary(naive.6)

```


# entry-level mixed effect modeling
```{r}
# unconditional means models

model.1 = lmer(exp_runs_change ~ 1 + (1|pitcher_id), data = saber1)
summary(model.1)
  # 1.119e-13 / (1.119e-13 + 1.248e-01) = 8.966346e-13
  # a tiny amount of the variation in changes in expected runs is attributable to differences         between pitchers compared to changes over time within each pitcher

  # average correlation between any pair of responses between pitchers is also tiny


model.1.2 = lmer(exp_runs_change ~ 1 + (1|g_id) + (1|pitcher_id), data = saber1)
summary(model.1.2)
  # 1.520e-13 / (1.520e-13 + 1.248e-01) = 1.217949e-12
  # still a very small amount of the variation is attributable to differences between pitchers        compared to the variation within pitchers and games, but higher than the 2 level model, exploring 3 level models is supported
  
```

```{r}
# unconditional growth models

model.2 = lmer(exp_runs_change ~ pitch_count + (pitch_count | pitcher_id), data = saber1)
summary(model.2)
  # residual var = 1.247e-01
  # residual var from uncond means = 1.248e-01
      # pseudo-R^2 : (1.248e-01 - 1.247e-01)/1.248e-01 = 0.0008012821

  # estimated within-pitcher variance dropped 0.08% when accounting for a linear increase in         pitch count
    # in other words : 0.08% of variability can be explained by a linear increase over time           (represented by pitch_count)


model.2.2 = lmer(exp_runs_change ~ pitch_count + (pitch_count | g_id) + (pitch_count | pitcher_id), data = saber1)


summary(model.2.2)
  # residual var = 1.246e-01
  # residual var from uncond means = 1.248e-01
    # pseudo-R^2 : (1.248e-01 - 1.246e-01)/1.248e-01 = 0.001602564

  # estimated within-pitcher variance dropped 0.16% when accounting for a linear increase in          pitch count and correlation within games
    # in other words: 0.16% of variability can be explained by a linear increase over time              (represented by pitch_count)


```

```{r}
# unconditional quadratic growth model

model.3 = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + (1|pitcher_id), data = saber1)
summary(model.3)

  # t values are all larger than 1.96 and all variables are significant
  
  # residual var from uncond quad growth = 1.246e-01
  # residual var from uncond means = 1.248e-01
    # pseudo R^2: (1.248e-01 - 1.246e-01) / 1.248e-01 = 0.001602564

  # estimated within-pitcher variance dropped 0.16% from the unconditional means model when           accounting for a quadratic increase in pitch count
    # in other words: 0.16% of variability can be explained by a quadratic increase in pitch            count over time


model.3.2 = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + (pitch_count | g_id) + (pitch_count | pitcher_id), data = saber1)

summary(model.3.2)
  # residual var = 1.245e-01
  # residual var from uncond means = 1.248e-01
      # pseudo R^2: (1.248e-01 - 1.245e-01)/1.248e-01 = 0.002403846

  # estimated within-pitcher variance dropped 0.24% from the unconditional means model when          accounting for a quadratic increase in pitch count
    # in other words:  0.24% of variability can be explained by a quadratic increase in pitch           count over time
```

```{r}

model.4 = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + top + game_state + (1|pitcher_id), data = saber1)
  # only game info without scores
summary(model.4)


model.5 = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + top + game_state + p_score + b_score + prev_ab_pc + (1|pitcher_id/g_id), data = saber1)
  # added game info (scores and prev_ab_pc)
summary(model.5)


model.6 = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + p_score + b_score + prev_ab_pc + (1|pitcher_id), data = saber1)
  # added interaction term for stand:p_throws (platoon advantage)
summary(model.6)


model.7= lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + p_score + b_score + prev_ab_pc + (1|pitcher_id/g_id), data = saber1)
  # nested g_id inside of p_id
summary(model.7)

model.8 = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + (1|pitcher_id), data = saber1)
  # removed insignificant variables scores and prev_ab_pc and disregarded nested games
summary(model.8)


model.9 = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + (pitch_count|pitcher_id), data = saber1)
  # random slope and intercept
summary(model.9)

```

### with smaller number of pitchers
```{r}
library(tidyverse)
freqvec = table(saber1$pitcher_id)
idvec <- names(freqvec[freqvec >= 2750])
sabersmall <- saber1 %>% filter(pitcher_id %in% idvec)
sabersmall$pitcher_id = as.character(as.factor(sabersmall$pitcher_id))


# unconditional means
model.1.s = lmer(exp_runs_change ~ 1 + (1|pitcher_id), data = sabersmall)
summary(model.1.s)
  # singular
  # 0/0.1225 = 0 variance explained

# unconditional means 3 level
model.1.2.s = lmer(exp_runs_change ~ 1 + (1|g_id) + (1|pitcher_id), data = sabersmall)
summary(model.1.2.s)
  # singular
  # 0 + 0 / 0.1225 = 0 variance explained

# unconditional growth
model.2.s = lmer(exp_runs_change ~ pitch_count + (pitch_count | pitcher_id), data = sabersmall)
summary(model.2.s)
  # singular
  # residual var =  1.224e-01
  # residual var from uncond. means 2 level = 0.1225
      # increase in explained variance: (0.1225 - 0.1224) / 0.1225 = 0.0008163265, 0.08%

# unconditional growth 3 level
model.2.2.s = lmer(exp_runs_change ~ pitch_count + (pitch_count | g_id) + (pitch_count | pitcher_id), data = saber1)
summary(model.2.2.s)
  # singular
  # residual var =  1.246e-01
  # residual var from uncond. means 3 level = 0.1225
      # increase in explained variance: (0.1225 - 0.1246) / 0.1225 = -0.017, decreased explained variance

# unconditional quadratic growth
model.3.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + (1|pitcher_id), data = sabersmall)
summary(model.3.s)
  #singular
  # residual var =  0.1222 
  # residual var from uncond. means 2 level = 0.1225
    # increase in explained variance: (0.1225 - 0.1222)  / 0.1225 = 0.002449, 0.2% increase

# unconditional quadratic growth 3 level
model.3.2.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + (pitch_count | g_id) + (pitch_count | pitcher_id), data = sabersmall)
summary(model.3.2.s)
  # singular
  # residual var = 0.1221
  # residual var from uncond. means = 0.1225
    # increase in explained variance: (0.1225 - 0.1221) / 0.1225 = 0.00327, 0.3% increase
```

```{r}
# advanced modeling

model.4.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + top + game_state + (1|pitcher_id), data = sabersmall)
  # only game info without scores
summary(model.4.s)


model.5.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + top + game_state + p_score + b_score + prev_ab_pc + (1|pitcher_id/g_id), data = sabersmall)
  # added game info (scores and prev_ab_pc)
summary(model.5.s)


model.6.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + p_score + b_score + prev_ab_pc + (1|pitcher_id), data = sabersmall)
  # added interaction term for stand:p_throws (platoon advantage)
summary(model.6.s)


model.7.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + p_score + b_score + prev_ab_pc + (1|pitcher_id/g_id), data = sabersmall)
  # nested g_id inside of p_id
summary(model.7.s)


model.8.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + prev_ab_pc + stand:p_throws + (1|pitcher_id), data = sabersmall)
  #disregarded nested games and scores
summary(model.8.s)

### FINAL MODEL
model.9.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + (1|pitcher_id), data = sabersmall)
  # disregarded prev_ab_pc
summary(model.9.s)

model.10.s = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + (pitch_count|pitcher_id), data = sabersmall)
  # random slope and intercept
summary(model.10.s)
```

# comparing reduced starter models
```{r}
AIC(model.1.2.s, model.2.2.s, model.3.2.s, model.4.s,  model.5.s, model.6.s, model.7.s, model.8.s,  model.9.s, model.10.s)
```


# mixed effect modeling for relievers
```{r}
saber2$inning = as.integer(saber2$inning)

model.4.r = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + top + game_state + (1|pitcher_id), data = saber2)
  # only game info without scores
summary(model.4.r)


model.5.r = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + top + game_state + p_score + b_score + prev_ab_pc + (1|pitcher_id/g_id), data = saber2)
  # added game info (scores and prev_ab_pc)
summary(model.5.r)


model.6.r = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + p_score + b_score + prev_ab_pc + (1|pitcher_id), data = saber2)
  # added interaction term for stand:p_throws (platoon advantage)
summary(model.6.r)


model.7.r = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + times_faced + stand:p_throws + p_score + b_score + prev_ab_pc + (pitch_count|pitcher_id), data = saber2)
  # added random slope


### FINAL MODEL
model.8.r = lmer(exp_runs_change ~ pitch_count + I(pitch_count^2) + inning + game_state + top + stand:p_throws + p_score + b_score + (1|pitcher_id), data = saber2)
  # removed random slope, times_faced and prev_ab_pc
summary(model.8.r)

```

# comparing reliever models
```{r}
AIC(model.4.r, model.5.r, model.6.r, model.7.r, model.8.r, model.9.r)
```

# model use
```{r}

# compare josh tomlin to andrew miller (cleveland guardians, 2016 season)

findplayer = player_names %>% filter(last_name =='Miller') %>% print()

inning = as.integer(6)
top = as.factor('False')
game_state = as.factor('0020')
stand = as.factor('R')
p_score = as.integer(0)
b_score = as.integer(2)

p_throws = as.factor('R') # starter
pitcher_id = as.factor('458708') # starter
times_faced =  3 # starter
pitch_count = 75 # starter

r_throws = as.factor('L') # reliever
reliever_id = as.factor('453192') # reliever
reliever_count = 0


starter_pred = data.frame(inning, top, game_state, stand, pitch_count, times_faced, p_throws, pitcher_id)
print('starter: josh tomlin')
predict(model.8, newdata = starter_pred)

reliever_pred = data.frame(inning, top, game_state, stand, r_throws, reliever_id, p_score, b_score, reliever_count) %>% rename(p_throws = r_throws, pitcher_id = reliever_id, pitch_count = reliever_count)
print('reliever: andrew miller')
predict(model.8.r, newdata = reliever_pred)

```

